Value-Suppressing Uncertainty Maps

### 2017-09-17 paper/main.tex
First round of JH changes

VSUPs rely on an underlying \textbf{quantization tree} that governs how values are discretized. Maximally uncertain values are mapped to a singular ``root'' node. As uncertainty decreases, the tree branches into leaves, which evenly divide the data domain. The data value of a parent is the midpoint of all of its children. These leaves can then branch again, up to a designer-specified stopping point. The layers quantize the uncertainty domain, and sibling nodes quantize the data domain. \figref{fig:example} shows a VSUP with a tree with a branching factor of 2 and 4 layers, resulting in an output range with 15 distinct colors.

Since the visual channel used to encode uncertainty (e.g., saturation, lightness, transparency) often reduces the ability of people to distinguish colors, VSUPs have the added benefit of creating bivariate palettes that are perceptually easier to distinguish, by reducing color resolution in precisely the areas where color disambiguation would be most difficult anyway. For instance, the two closest colors in the VSUP in \figref{fig:example} are 1.4 units farther apart in CIELAB color space.

JavaScript code for generating VSUPs for use in D3 charts is available at \url{URL-REMOVED-FOR-REVIEW}.

@@

When selecting a visual channel to represent value, the designer should consider the perceptual integrality of the value and uncertainty visual channels. For instance, we use the Viridis~\cite{viridis} color map frequently in this paper, as many other standard color ramps (such as sequential ColorBrewer~\cite{harrower2003colorbrewer} ramps), following usual best practices, interpolate in both hue and value~\cite{ware1988color}. This interpolation in value interferes with our uncertainty encoding, introducing ambiguity. Ramps with similar saturation and value among colors, such as Viridis, are less ambiguous.

When determining the quantization tree, the design should be careful to avoid having too many output colors. For reasons of memorability and complexity, Wainer\& Francolini~\cite{wainer1980empirical} suggest no more than 16. For a binary quantization tree, this results in a maximum tree depth of 4, as seen in \figref{fig:example}. In general, for tasks that require finer-grade distinctions between uncertainty, the depth of the tree should be increased. For tasks involving finer-grade distinctions between value, the branching factor of the tree should be increased. Beyond these two parameters, non-uniform binning would allow the designer to target particular distributions or important subregions of the data that would otherwise be difficult to visualize in a limited palette, as in Dunn's~\cite{dunn1989dynamic} dynamic bivariate maps. As an additional form of control, prior tools using VSUP-like techniques, such as LayerCake~\cite{correll2015layercake}, allow the analyst to interactively control the size of bins in the tree through dragging on the color legend.

\subsection{Examples}

We present a set of examples showing how VSUPs can be applied to real-world datasets. The decision of VSUPs to allocate color bins asymmetrically amongst value and uncertainty regions affords greater discriminability in regions of the chart that warrant closer scrutiny, while discouraging exploration of regions with noisy or unreliable signals.

\subsubsection{Air Travel Delay}

@@

\subsubsection{Viral Mutation}
%TODO update this figure
\viralFig

@@

Rather than standard means, Cleveland \& McGill~\cite{cleveland1984graphical} use 25\% trimmed means to report effect sizes in graphical perception tasks. Trimmed means remove the first and last quartiles and then compute the mean from the remaining points. This filtering is important for graphical perception tasks (especially crowdsourced tasks) where error distributions may be long-tailed~\cite{heer2010crowdsourcing}. Trimmed means violate the sampling assumptions required for many null-hypothesis significance tests. Thus, while we report effect sizes in terms of trimmed means, we perform inferential statistical tests on the standard means. Asterisks in our charts denote significant differences, as reported by a post-hoc Tukey's test of Honest Significant Difference (HSD).

@@

For the identification task, we gave participants a $5$x$5$ heatmap, and asked them to click on a region of the heatmap that encodes a particular (value, uncertainty) pair. To support comparability across conditions, we chose a set of target pairs such that each target mapped to a unique color across all encoding types. This resulted in (4+4+2+1)=11 valid targets. For each trial, these 11 targets were placed randomly in a heatmap; the remaining 13 cells of the heatmap were distractors randomly sampled from this target list.

@@

The results support our first hypothesis. Juxtaposition was a significant effect ($F(1,166)=5.5$,$p=0.02$). A post-hoc t-test confirmed ($t=2.5$, $p=0.01$) that superimposed charts ($M=0.58$, $SD=0.5$) performed significantly better than juxtaposed ($M=0.51$, $SD=0.49$) ones. For our identification task, juxtaposition requires at least two search tasks (one to locate squares with the correct data value, another to locate ones with the correct uncertainty value), which introduces additional error compared to conditions where value and uncertainty are displayed simultaneously.

@@

Our results from the prior study indicate that discrete, non-juxtaposed maps outperformed the other bivariate maps we selected, so we limited our study to only 4 types: square and arc bivariate maps. These in turn were either VSUPs or traditional bivariate maps, for a 2 (square or arc legend) x 2 (VSUP or standard) factorial design, with 6 replications, for a total of 24 stimuli for each of our 24 participants. \figref{fig:conditions} shows these factor levels. Prior to the main task, we included a short replication of the identification task from the prior experiment (with 12 stimuli) for training purposes. Poor performance on this training was used for exclusion purposes. Three people with low accuracy ($6\%$, $6\%$, and $25\%$ accuracy compared to a mean of $70\%$) were excluded.

@@

Bivariate maps have an inherent limitation in that visual channels are often difficult to attend to separately or orthogonally. Bivariate maps are also complex to interpret. Systems which simultaneously display large amounts of value and uncertainty data to general audiences (such as Pangloss~\cite{moritz2017trust}) avoid them for precisely these reasons, relying instead on juxtaposed maps. Our results indicate that juxtaposition produces additional problems, forcing users that wish to integrate uncertainty and data to locate an item of interest, and then re-locate that same item in a visually distant map, which may have few visual landmarks in common. This additional step before information fusion is reflected in longer response times, and patterns of decision-making that are less mindful of risk. Given the deleterious effects of latency on data analysis~\cite{liu2014effects}, designers ought to carefully consider the tradeoffs between expressiveness of visual encoding, and usability.

>>>>

VSUPs rely on an underlying \textbf{quantization tree} that governs how values are discretized. Values about a certain uncertainty threshold are mapped to a singular ``root'' node. As uncertainty decreases, the tree branches into leaves, which evenly divide the data domain. The data value of a parent is the midpoint of all of its children. These leaves can then branch again, up to a designer-specified stopping point. The layers quantize the uncertainty domain, and sibling nodes quantize the data domain. \figref{fig:example} shows a VSUP with a tree with a branching factor of 2 and 4 layers, resulting in an output range with 15 distinct colors.

Since the visual channel used to encode uncertainty (e.g., saturation, lightness, transparency) often reduces the ability of people to distinguish colors, VSUPs have the added benefit of creating bivariate palettes that are perceptually easier to distinguish, by reducing color resolution in precisely the areas where color disambiguation would be most difficult anyway. For instance, the two closest colors in the VSUP in \figref{fig:example} are 1.4 units farther apart in CIELAB color space than the two closest colors in the traditional bivariate map.

JavaScript code for generating VSUPs for use in D3~\cite{bostock2011d3} charts is available at \url{URL-REMOVED-FOR-REVIEW}.

@@

When selecting a visual channel to represent value, the designer should consider the perceptual integrality of the value and uncertainty visual channels. For instance, we use the Viridis~\cite{viridis} color map frequently in this paper, as many other standard color ramps (such as sequential ColorBrewer~\cite{harrower2003colorbrewer} ramps), following usual best practices, interpolate in both hue and luminance~\cite{ware1988color}. This interpolation in value interferes with our uncertainty encoding, introducing ambiguity. Ramps such as Viridis, which avoid very light and very dark colors, reduce this ambiguity.

When determining the quantization tree, the design should be careful to avoid having too many output colors. For reasons of memorability and complexity, Wainer\& Francolini~\cite{wainer1980empirical} suggest no more than 16. For a binary quantization tree, this results in a maximum tree depth of 4, as seen in \figref{fig:example}. In general, for tasks that require finer-grade distinctions between uncertainty, the depth of the tree could be increased. For tasks involving finer-grade distinctions between value, the branching factor of the tree could be increased. However, these factors increase the number of colors, and so designers should moderate between legend complexity and data fidelity. Beyond these two parameters, non-uniform binning would allow the designer to target particular distributions or important subregions of the data that would otherwise be difficult to visualize in a limited palette, as in Dunn's~\cite{dunn1989dynamic} dynamic bivariate maps. As an additional form of control, prior tools using VSUP-like techniques, such as LayerCake~\cite{correll2015layercake}, allow the analyst to interactively control the size of bins in the tree through dragging on the color legend.

\section{Examples}

We present a set of examples showing how VSUPs can be applied to real-world datasets. VSUPs allocate color bins asymmetrically amongst value and uncertainty regions, affording greater discriminability in regions of the chart that warrant closer scrutiny, while discouraging exploration of regions with noisy or unreliable signals.

\subsection{Air Travel Delay}

@@

\subsection{Viral Mutation}
%TODO update this figure
\viralFig

@@

%Rather than standard means, Cleveland \& McGill~\cite{cleveland1984graphical} use 25\% trimmed means to report effect sizes in graphical perception tasks. Trimmed means remove the first and last quartiles and then compute the mean from the remaining points. This filtering is important for graphical perception tasks (especially crowdsourced tasks) where error distributions may be long-tailed~\cite{heer2010crowdsourcing}. Trimmed means violate the sampling assumptions required for many null-hypothesis significance tests. Thus, while we report effect sizes in terms of trimmed means, we perform inferential statistical tests on the standard means. Asterisks in our charts denote significant differences, as reported by a post-hoc Tukey's test of Honest Significant Difference (HSD).

@@

For the identification task, we gave participants a $5$x$5$ heatmap, and asked them to click on a region of the heatmap that encodes a particular (value, uncertainty) pair. To support comparability across conditions, we chose a set of target pairs such that each target mapped to a unique color across all encoding types. This resulted in (4+4+2+1)=11 valid targets. For each trial, we placed one example of each of these 11 targets randomly in a heatmap; the remaining 13 cells of the heatmap were randomly sampled from this target list.

@@

The results support our first hypothesis. Juxtaposition was a significant effect ($F(1,166)=5.5$,$p=0.02$). A post-hoc t-test confirmed ($t=2.5$, $p=0.01$) that superimposed charts ($M=0.58$, $SD=0.5$) performed significantly better than juxtaposed ($M=0.51$, $SD=0.49$) ones. For our identification task, juxtaposition requires at least two search tasks (one to locate squares with the correct data value, another to locate ones with the correct uncertainty value), which introduces additional error compared to conditions where value and uncertainty are displayed simultaneously. This is reflected also in response times, which, although highly variable, were highest in juxtaposed conditions ($M=19$, $SD=65$), than the other conditions ($M=13$, $SD=38$), although this difference was not significant ($t=1.64$, $p=0.1$).

@@

Our results from the prior study indicate that discrete, non-juxtaposed maps outperformed the other bivariate maps we selected, so we limited our study to only 4 types: square and wedge bivariate maps. These in turn were either VSUPs or traditional bivariate maps, for a 2 (square or wedge legend) x 2 (VSUP or standard) factorial design, with 6 replications, for a total of 24 stimuli for each of our 24 participants. \figref{fig:conditions} shows these factor levels. Prior to the main task, we included a short replication of the identification task from the prior experiment (with 12 stimuli) for training purposes. Poor performance on this training was used for exclusion purposes. Three people with low accuracy ($6\%$, $6\%$, and $25\%$ accuracy compared to a mean of $70\%$) were excluded.

@@

Bivariate maps have an inherent limitation in that visual channels are often difficult to attend to separately or orthogonally. Bivariate maps are also complex to interpret. Systems which simultaneously display large amounts of value and uncertainty data to general audiences (such as Pangloss~\cite{moritz2017trust}) avoid them for precisely these reasons, relying instead on juxtaposed maps. Our results indicate that juxtaposition produces additional problems, forcing users that wish to integrate uncertainty and data to locate an item of interest, and then re-locate that same item in a visually distant map, which may have few visual landmarks in common. This additional step before information fusion is reflected in longer response times, and patterns of decision-making that are less mindful of risk.
